RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing

### 2013-09-02 doc/bioinformatics2013/RVD2.tex
edited methods.

\title{RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing}
\author{}
%\date{}                                           % Activate to display a given date or no date

@@

Instead of attempting to directly compute the posterior distribution, we instead sample it using a Metropolis-Hastings within Gibbs scheme as described in Metropolis-within-Gibbs section.

\section{Inference and Hypthesis Testing}
\subsection{Model parameter estimation}\

Method of moments is used to find the optimal model parameters and hyperparameters. From the feature of beta distribution
The estimate of

@@

\subsection{Metropolis-within-Gibbs sampling}\

Because of the computation complexities of the normalization factor in the posterior probability, A Markov Chain Monte Carlo (MCMC) sampling approach is employed to approximate the posterior distribution $p(\theta, \mu | r,n; \mu_0,M_0,M)$. Specifically Metropolis-within-Gibbs algorithm is used to obtain the posterior distribution of each latent variable.

\begin{algorithm}[ht]
\caption{Metropolis within Gibbs Algorithm}
\label{alg:metro_gibbs}
\begin{algorithmic}[1]
\Repeat
\For {j = 1 to J}
  \For {i =1 to N}
	\State Draw 1 samples from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$ using Gibbs; sampling
  \EndFor
\EndFor

\For {j = 1 to J}
	\State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H;
	\State Set $mu_j$ to the sample median;
\EndFor
\Until {sample size is sufficient}

\end{algorithmic}
\end{algorithm}

\subsection{Statistic Inference}\

Full Bayesian Hypothesis Testing is done to test rare variants in the model.

\subsubsection{Bayesian Hypothesis Testing}
To detect a mutation in a specific position, a threshold $\tau$ is set for the posterior difference distribution testing.
\begin{align}
 H_1: \mu_{case}-\mu_{control}\geq\tau \\
 H_2: \mu_{case}-\mu_{control}<\tau
\end{align}

@@

\subsubsection{Goodness of fit Testing for Uniformity}\

@@

\section{Result}
Figure 1 Overall of the process
\subsection{Simulation Results}
\subsection{Comparison Results on Synthetic DNA}
Figure 3 ROC comparison with other methods: varscan/samtools/rvd
\subsection{Empirical Results on Multiple Sclerosis Data}
We tested our method on sequence data from clinical Multiple Sclerosis samples. We expect to see that

>>>>

\algnewcommand{\Inputs}[1]{%
  \State \textbf{Inputs:}
  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
}
\algnewcommand{\Initialize}[1]{%
  \State \textbf{Initialize:}
  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
}

\title[RVD2]{RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing}
\author{}
%\date{}                                           % Activate to display a given date or no date

@@

Instead of attempting to directly compute the posterior distribution, we sample it using a Metropolis-Hastings within Gibbs scheme as described in Section~\ref{sec:algorithm}.

\section{Inference and Hypothesis Testing}

The primary object of inference in this model the joint posterior distribution function over the latent variables,
\begin{equation}
	p(\mu, \theta | r, n; \phi)  = \frac{ p(\mu, \theta, r | n; \phi) } {p ( r | n; \phi)},
\end{equation}
where the parameters are $\phi \triangleq \{\mu_0, M_0, M\}$.

We use a Metropolis-within-Gibbs sampling approach shown in Algorithm~\ref{alg:metro_gibbs} to approximate the posterior distribution. The model parameters are initialized using method-of-moments. Given those parameter estimates we first sample from the marginal posterior distribution for $\mu_j$ given it's Markov blanket using Metropolis Hasting. Then, we sample from the marginal posterior distribution for $\theta_{ji}$ given it's Markov blanket.

\begin{algorithm}[ht]
\caption{Metropolis-within-Gibbs Algorithm}
\label{alg:metro_gibbs}
\begin{algorithmic}[1]

\State Initialize $\theta$, $\mu$, $M$, $\mu_0$, $M_0$
\Repeat
\For {each position j} \Comment{Sample $\mu_j$}
  \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
  \State Set $\mu_j$ to the sample median for the T samples

  \For {each replicate i} \Comment{Sample $\theta_{ji}$}
	\State Sample from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$
  \EndFor

\EndFor
\Until {convergence and sample size sufficient}
\end{algorithmic}
\end{algorithm}

Once we have obtained an estimate of the posterior distribution, we test the hypothesis that there is a sub-population of DNA molecules in the pool with a non-reference base at position $j, \forall j \in 1, \ldots, J$. First, we test the hypothesis that the error rate in the case data, $\mu_j^{\text{case}}$ is greater than the error rate in the control data, $\mu_j^{\text{control}}$. If so, then we test the hypothesis that the high error rate is due to excess reads with a particular non-reference base rather than a uniform distribution over all possible non-reference bases.

\subsection{Initialization}

The initial values of the model parameters and latent variables is obtained by a method-of-moments procedure. Essentially, we set the parameter value (population moment) equal to the sample moment. In some cases, the method-of-moments estimate is outside the parameter space. Then, we XXX.

@@

\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}

We are able to draw samples from $p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$ analytically because of the Bayesian conjugacy between the prior $p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$ and the likelihood $p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.

\subsection{Sampling from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$}

\subsection{Bayesian hypothesis test for high error rate}

To detect a mutation in a specific position, a threshold $\tau$ is set for the posterior difference distribution testing.
\begin{align}
 H_1: \mu_{case}-\mu_{control}\geq\tau \\
 H_2: \mu_{case}-\mu_{control}<\tau
\end{align}

@@

\subsection{$\chi^2$ test for non-uniform base distribution}

@@

\section{Results}
Figure 1 Overall of the process
\subsection{Simulation Results}
\subsection{Comparison Results on Synthetic DNA}
Figure 3 ROC comparison with other methods: varscan/samtools/rvd
\subsection{Empirical Results on Multiple Sclerosis Data}
We tested our method on sequence data from clinical Multiple Sclerosis samples. We expect to see that
